---
title: "Prediction of 30-Day Hospital Readmission in Patients with Diabetes"
author: "Sara Raees"
output: html_document
date: "2026-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(skimr)
library(caret)
library(naniar)
library(pROC)
library(randomForest)
```


#Background

This project uses the Diabetes 130-US Hospitals for Years 1999–2008 dataset from the UCI Machine Learning Repository. The aim is to develop a prediction model for 30-day hospital readmission among hospitalized patients with diabetes.

#Load, Read and Inspect the dataset
```{r}
diabetes <- read.csv("diabetic_data.csv", stringsAsFactors = FALSE)

dim(diabetes)

names(diabetes)

```
#Initital data overview
```{r}
glimpse(diabetes)
summary(diabetes)
```

#Replace "?" with NA

```{r}
diabetes <- diabetes %>%
  mutate(across(where(is.character), ~ na_if(., "?")))
```


#Define the outcome variable
The original variable readmitted has three categories:

"NO" = no readmission
">30" = readmission after 30 days
"<30" = readmission within 30 days

We create a binary outcome:

1 = readmitted within 30 days
0 = not readmitted within 30 days

```{r}
diabetes <- diabetes %>%
  mutate(readmit30 = ifelse(readmitted == "<30", 1, 0),
         readmit30 = factor(readmit30, levels = c(0, 1),
                            labels = c("No", "Yes")))
```

#Study population and outcome frequency
```{r}
# Total number of observations
nrow(diabetes)

# Total number of predictors before excluding IDs/outcome
ncol(diabetes) - 1

# Outcome distribution
table(diabetes$readmitted)
table(diabetes$readmit30)

# Proportion readmitted within 30 days
prop.table(table(diabetes$readmit30))
```
#Count number of patients readmitted within 30 days
```{r}
sum(diabetes$readmitted == "<30", na.rm = TRUE)
```


#Missing data assessment
```{r}
# Number of missing values per variable
missing_counts <- colSums(is.na(diabetes))

# Proportion missing per variable
missing_prop <- colMeans(is.na(diabetes)) * 100

missing_summary <- data.frame(
  Variable = names(diabetes),
  Missing_Count = missing_counts,
  Missing_Percent = round(missing_prop, 2)
) %>%
  arrange(desc(Missing_Percent))

missing_summary
```


```{r}
# Visualize missingness
diabetes %>%
  slice_sample(n = 5000) %>%
  vis_miss()
```

#Descriptive analysis of selected variables
```{r}
# Age distribution
table(diabetes$age)

# Gender distribution
table(diabetes$gender, useNA = "ifany")

# Race distribution
table(diabetes$race, useNA = "ifany")

# Summary of selected numeric variables
diabetes %>%
  select(time_in_hospital, num_lab_procedures, num_procedures,
         num_medications, number_outpatient, number_emergency,
         number_inpatient, number_diagnoses) %>%
  summary()
```

#Select variables for modelling

Remove identifier variables that should not be used as predictors.
```{r}
diabetes_model <- diabetes %>%
  select(-encounter_id, -patient_nbr, -readmitted)
```

#Convert character variables to factors
```{r}
diabetes_model <- diabetes_model %>%
  mutate(across(where(is.character), as.factor))
```

#Remove variables with very high missingness

```{r}
missing_threshold <- 0.5

diabetes_model <- diabetes_model %>%
  select(where(~ mean(is.na(.)) < missing_threshold))

dim(diabetes_model)
```

#Train/test split
```{r}
set.seed(123)

train_index <- createDataPartition(diabetes_model$readmit30, p = 0.7, list = FALSE)

train_data <- diabetes_model[train_index, ]
test_data  <- diabetes_model[-train_index, ]

dim(train_data)
dim(test_data)
```

#Simple imputation

For a first analysis:

- numeric variables: median imputation
- categorical variables: create "Missing" level
```{r}
train_data_imputed <- train_data
test_data_imputed  <- test_data

# Numeric median imputation
num_vars <- names(train_data_imputed)[sapply(train_data_imputed, is.numeric)]

for (v in num_vars) {
  med <- median(train_data_imputed[[v]], na.rm = TRUE)
  train_data_imputed[[v]][is.na(train_data_imputed[[v]])] <- med
  test_data_imputed[[v]][is.na(test_data_imputed[[v]])] <- med
}

# Categorical missing category
cat_vars <- names(train_data_imputed)[sapply(train_data_imputed, is.factor)]

for (v in cat_vars) {
  train_data_imputed[[v]] <- addNA(train_data_imputed[[v]])
  levels(train_data_imputed[[v]])[is.na(levels(train_data_imputed[[v]]))] <- "Missing"
  
  test_data_imputed[[v]] <- factor(test_data_imputed[[v]],
                                   levels = levels(train_data_imputed[[v]]))
  test_data_imputed[[v]] <- addNA(test_data_imputed[[v]])
  levels(test_data_imputed[[v]])[is.na(levels(test_data_imputed[[v]]))] <- "Missing"
}
```

#Logistic regression model
```{r}
# Convert character to factor
train_data_imputed <- train_data_imputed %>%
  mutate(across(where(is.character), as.factor))

test_data_imputed <- test_data_imputed %>%
  mutate(across(where(is.character), as.factor))

# Match test levels to train
for (v in names(train_data_imputed)) {
  if (is.factor(train_data_imputed[[v]]) && v %in% names(test_data_imputed)) {
    test_data_imputed[[v]] <- factor(test_data_imputed[[v]],
                                     levels = levels(train_data_imputed[[v]]))
  }
}

# Drop unused levels
train_data_imputed <- train_data_imputed %>%
  mutate(across(where(is.factor), droplevels))

test_data_imputed <- test_data_imputed %>%
  mutate(across(where(is.factor), droplevels))

# Remove factor columns with fewer than 2 observed categories
bad_vars <- names(train_data_imputed)[sapply(train_data_imputed, function(x) {
  if (is.factor(x)) {
    length(unique(na.omit(x))) < 2
  } else {
    FALSE
  }
})]

bad_vars

train_data_imputed <- train_data_imputed %>% select(-all_of(bad_vars))
test_data_imputed  <- test_data_imputed %>% select(-all_of(bad_vars))

# Fit logistic regression
logit_model <- glm(readmit30 ~ ., data = train_data_imputed, family = binomial)

summary(logit_model)
```

#Test set predictions

```{r}
pred_prob <- predict(logit_model, newdata = test_data_imputed, type = "response")
pred_class <- ifelse(pred_prob > 0.5, "Yes", "No") %>% factor(levels = c("No", "Yes"))

confusionMatrix(pred_class, test_data_imputed$readmit30, positive = "Yes")
```
```{r}
# Predicted probabilities on test data
logit_pred_prob <- predict(logit_model, newdata = test_data_imputed, type = "response")

# Predicted classes using 0.5 cutoff
logit_pred_class <- ifelse(logit_pred_prob > 0.5, "Yes", "No") %>%
  factor(levels = c("No", "Yes"))

# Confusion matrix
confusionMatrix(logit_pred_class, test_data_imputed$readmit30, positive = "Yes")
```
```{r}
# ROC and AUC
roc_logit <- roc(response = test_data_imputed$readmit30,
                 predictor = logit_pred_prob,
                 levels = c("No", "Yes"))

auc(roc_logit)
plot(roc_logit, main = "ROC Curve - Logistic Regression")
```
```{r}
# Create calibration data
calibration_df <- data.frame(
  observed = ifelse(test_data_imputed$readmit30 == "Yes", 1, 0),
  predicted = logit_pred_prob
)

# Group predictions into deciles
calibration_plot_data <- calibration_df %>%
  mutate(decile = ntile(predicted, 10)) %>%
  group_by(decile) %>%
  summarise(
    mean_predicted = mean(predicted),
    mean_observed = mean(observed),
    .groups = "drop"
  )

calibration_plot_data
```

```{r}
# Calibration plot
ggplot(calibration_plot_data, aes(x = mean_predicted, y = mean_observed)) +
  geom_point(size = 3) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    title = "Calibration Plot - Logistic Regression",
    x = "Mean Predicted Probability",
    y = "Observed Proportion"
  ) +
  theme_minimal()
```
```{r}
# Brier score
brier_logit <- mean((calibration_df$observed - calibration_df$predicted)^2)
brier_logit
```

```{r}
train_rf <- train_data_imputed %>%
  select(-any_of(c("diag_1", "diag_2", "diag_3", "medical_specialty")))

test_rf <- test_data_imputed %>%
  select(-any_of(c("diag_1", "diag_2", "diag_3", "medical_specialty")))
```


```{r}
set.seed(123)

rf_model <- randomForest(
  readmit30 ~ .,
  data = train_rf,
  ntree = 300,
  importance = TRUE
)

print(rf_model)
```

```{r}
rf_pred_class <- predict(rf_model, newdata = test_rf, type = "class")
rf_pred_prob  <- predict(rf_model, newdata = test_rf, type = "prob")[, "Yes"]

confusionMatrix(rf_pred_class, test_rf$readmit30, positive = "Yes")
```


#Brief summary
```{r}
cat("Total observations:", nrow(diabetes), "\n")
cat("Total variables:", ncol(diabetes), "\n")
cat("Number readmitted within 30 days:", sum(diabetes$readmitted == "<30", na.rm = TRUE), "\n")
cat("Proportion readmitted within 30 days:",
    round(mean(diabetes$readmitted == "<30", na.rm = TRUE) * 100, 2), "%\n")
```

