---
title: "Prediction of 30-Day Hospital Readmission in Patients with Diabetes"
author: "Sara Raees"
date: "2026-03-29"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(janitor)
library(skimr)
library(caret)
library(naniar)
library(pROC)
library(randomForest)
library(dplyr)
library(scales)
```

## Background
- **Dataset**: Diabetes 130-US Hospitals for Years 1999–2008 (UCI Machine Learning Repository).
- **Aim**: Develop a prediction model for 30-day hospital readmission among hospitalized patients with diabetes.
- **Approach**: A stratified sample was used to preserve the proportion of patients readmitted within 30 days.

```{r data-load, include=FALSE, cache=TRUE}
diabetes_full <- read.csv("diabetic_data.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_index <- createDataPartition(diabetes_full$readmitted, p = 20000 / nrow(diabetes_full), list = FALSE)
diabetes <- diabetes_full[sample_index, ]
diabetes[diabetes == "?"] <- NA
diabetes <- diabetes %>%
  mutate(across(where(is.character), ~ trimws(.))) %>%
  mutate(across(where(is.character), ~ na_if(., "?"))) %>%
  mutate(
    readmit30 = ifelse(readmitted == "<30", 1, 0),
    readmit30 = factor(readmit30, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

## Study Population
- **Total observations**: `r nrow(diabetes)`
- **Proportion readmitted within 30 days**: `r round(mean(diabetes$readmitted == "<30", na.rm = TRUE) * 100, 2)`%
- **Outcome**: Binary variable `readmit30` ("Yes" / "No")

## Missingness Analysis

```{r missingness, fig.width=8, fig.height=5}
diabetes %>%
  slice_sample(n = min(5000, nrow(diabetes))) %>%
  vis_miss()
```

**Interpretation**: Most variables have near-complete data. However, variables such as weight, medical_specialty, and payer_code show high missingness. Predictors with >50% missing data are excluded prior to modeling.

## Analysis: Age Group

```{r age, fig.width=8, fig.height=5}
ggplot(diabetes, aes(x = age, fill = readmit30)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("No" = "#4DAF4A", "Yes" = "#E41A1C")) +
  labs(
    title = "Proportion of 30-Day Readmission by Age Group",
    x = "Age Group",
    y = "Proportion",
    fill = "Readmitted in 30 Days"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Interpretation**: Readmission proportion slightly varies across age groups. While older patients make up a large portion of the cohort, the proportion of those readmitted within 30 days is comparable across advanced age categories.

## Analysis: Time in Hospital

```{r hosp, fig.width=8, fig.height=4.5}
ggplot(diabetes, aes(x = readmit30, y = time_in_hospital, fill = readmit30)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("No" = "#377EB8", "Yes" = "#FF7F00")) +
  labs(
    title = "Time in Hospital by 30-Day Readmission Status",
    x = "30-Day Readmission",
    y = "Time in Hospital"
  ) +
  theme_minimal()
```

**Interpretation**: Patients readmitted within 30 days ("Yes") tend to have a slightly higher median time in the hospital during their initial stay compared to those not readmitted ("No"), suggesting longer initial stays may relate to higher readmission risk.

## Analysis: Number of Medications

```{r meds, fig.width=8, fig.height=4.5}
ggplot(diabetes, aes(x = num_medications, fill = readmit30)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("No" = "#984EA3", "Yes" = "#E41A1C")) +
  labs(
    title = "Distribution of Number of Medications",
    x = "Number of Medications",
    y = "Count",
    fill = "Readmitted in 30 Days"
  ) +
  theme_minimal()
```

**Interpretation**: The distribution of number of medications is right-skewed. The group readmitted within 30 days has a notably wide spread in the number of prescribed medications, which may be a proxy for more complex or severe underlying health conditions.

## Methods
- **Identifier Removal**: Excluded `encounter_id`, `patient_nbr`.
- **Threshold**: Removed predictors with >50% missingness.
- **Train/Test Split**: 70% Training / 30% Testing.
- **Imputation**:
  - Numeric: Median Imputation
  - Categorical: Added "Missing" category
- **Models Evaluated**:
  1. Logistic Regression
  2. Random Forest (`ntree = 300`)

```{r methods-prep, include=FALSE}
diabetes_model <- diabetes %>%
  select(-encounter_id, -patient_nbr, -readmitted) %>%
  mutate(across(where(is.character), as.factor))

missing_threshold <- 0.5
diabetes_model <- diabetes_model %>%
  select(where(~ mean(is.na(.)) < missing_threshold))

set.seed(123)
train_index <- createDataPartition(diabetes_model$readmit30, p = 0.7, list = FALSE)
train_data <- diabetes_model[train_index, ]
test_data  <- diabetes_model[-train_index, ]

train_data_imputed <- train_data
test_data_imputed  <- test_data

num_vars <- names(train_data_imputed)[sapply(train_data_imputed, is.numeric)]
for (v in num_vars) {
  med <- median(train_data_imputed[[v]], na.rm = TRUE)
  if (is.na(med)) med <- 0
  train_data_imputed[[v]][is.na(train_data_imputed[[v]])] <- med
  test_data_imputed[[v]][is.na(test_data_imputed[[v]])] <- med
}

cat_vars <- names(train_data_imputed)[sapply(train_data_imputed, is.factor)]
for (v in cat_vars) {
  train_data_imputed[[v]] <- addNA(train_data_imputed[[v]])
  levels(train_data_imputed[[v]])[is.na(levels(train_data_imputed[[v]]))] <- "Missing"
  test_data_imputed[[v]] <- factor(test_data_imputed[[v]], levels = levels(train_data_imputed[[v]]))
  test_data_imputed[[v]] <- addNA(test_data_imputed[[v]])
  levels(test_data_imputed[[v]])[is.na(levels(test_data_imputed[[v]]))] <- "Missing"
}

train_data_imputed <- train_data_imputed %>% mutate(across(where(is.character), as.factor)) %>% mutate(across(where(is.factor), droplevels))
test_data_imputed <- test_data_imputed %>% mutate(across(where(is.character), as.factor)) %>% mutate(across(where(is.factor), droplevels))

bad_vars <- names(train_data_imputed)[sapply(train_data_imputed, function(x) { if (is.factor(x)) length(unique(na.omit(x))) < 2 else FALSE })]
train_data_imputed <- train_data_imputed %>% select(-all_of(bad_vars))
test_data_imputed  <- test_data_imputed %>% select(-all_of(bad_vars))

nzv <- nearZeroVar(train_data_imputed)
if (length(nzv) > 0) {
  train_data_imputed <- train_data_imputed[, -nzv, drop = FALSE]
  test_data_imputed  <- test_data_imputed[, colnames(train_data_imputed), drop = FALSE]
}
```

```{r logit-model, include=FALSE}
logit_model <- glm(readmit30 ~ ., data = train_data_imputed, family = binomial)
test_data_imputed$readmit30 <- factor(test_data_imputed$readmit30, levels = c("No", "Yes"))
factor_cols_logit <- names(train_data_imputed)[sapply(train_data_imputed, is.factor)]
for (col in factor_cols_logit) {
  if (col %in% names(test_data_imputed)) {
    test_data_imputed[[col]] <- as.character(test_data_imputed[[col]])
    test_data_imputed[[col]][!(test_data_imputed[[col]] %in% levels(train_data_imputed[[col]]))] <- NA
    test_data_imputed[[col]] <- factor(test_data_imputed[[col]], levels = levels(train_data_imputed[[col]]))
  }
}
test_logit_clean <- test_data_imputed[complete.cases(test_data_imputed), ]
logit_pred_prob <- predict(logit_model, newdata = test_logit_clean, type = "response")
roc_logit <- roc(response = test_logit_clean$readmit30, predictor = logit_pred_prob, levels = c("No", "Yes"), direction = "<")
calibration_df <- data.frame(observed = ifelse(test_logit_clean$readmit30 == "Yes", 1, 0), predicted = logit_pred_prob)
calibration_plot_data <- calibration_df %>% mutate(decile = ntile(predicted, 10)) %>% group_by(decile) %>% summarise(mean_predicted = mean(predicted), mean_observed = mean(observed), .groups = "drop")
brier_logit <- mean((calibration_df$observed - calibration_df$predicted)^2)
```

## Logistic Regression: ROC Curve

```{r logit-roc, fig.width=8, fig.height=5.5}
plot(roc_logit, col = "#1F78B4", lwd = 3, main = "ROC Curve - Logistic Regression")
abline(a = 0, b = 1, lty = 2, col = "grey50")
```

## Logistic Regression: Calibration

```{r logit-calib, fig.width=8, fig.height=5}
ggplot(calibration_plot_data, aes(x = mean_predicted, y = mean_observed)) +
  geom_point(size = 3, colour = "#E41A1C") +
  geom_line(colour = "#E41A1C") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey40") +
  labs(
    title = "Calibration Plot - Logistic Regression",
    x = "Mean Predicted Probability",
    y = "Observed Proportion"
  ) +
  theme_minimal()
```

```{r rf-model, include=FALSE}
high_card_vars <- names(train_data_imputed)[sapply(train_data_imputed, function(x) { is.factor(x) && nlevels(x) > 53 })]
train_rf <- train_data_imputed %>% select(-all_of(high_card_vars))
test_rf <- test_data_imputed %>% select(all_of(names(train_rf)))
train_rf$readmit30 <- factor(train_rf$readmit30, levels = c("No", "Yes"))
test_rf$readmit30  <- factor(test_rf$readmit30,  levels = c("No", "Yes"))

set.seed(123)
rf_model <- randomForest(readmit30 ~ ., data = train_rf, ntree = 300, importance = TRUE)

factor_cols <- names(train_rf)[sapply(train_rf, is.factor)]
for (col in factor_cols) {
  if (col %in% names(test_rf)) {
    test_rf[[col]] <- as.character(test_rf[[col]])
    test_rf[[col]][!(test_rf[[col]] %in% levels(train_rf[[col]]))] <- NA
    test_rf[[col]] <- factor(test_rf[[col]], levels = levels(train_rf[[col]]))
  }
}
for (col in names(train_rf)) {
  if (col %in% names(test_rf)) {
    if (is.numeric(train_rf[[col]]) && !is.factor(train_rf[[col]])) test_rf[[col]] <- as.numeric(test_rf[[col]])
    else if (is.integer(train_rf[[col]])) test_rf[[col]] <- as.integer(test_rf[[col]])
    else if (is.logical(train_rf[[col]])) test_rf[[col]] <- as.logical(test_rf[[col]])
  }
}
test_rf_clean <- test_rf[complete.cases(test_rf), ]
rf_pred_class <- predict(rf_model, newdata = test_rf_clean, type = "class")
rf_pred_prob  <- predict(rf_model, newdata = test_rf_clean, type = "prob")[, "Yes"]
roc_rf <- roc(response = test_rf_clean$readmit30, predictor = rf_pred_prob, levels = c("No", "Yes"), direction = "<")
rf_calibration_df <- data.frame(observed = ifelse(test_rf_clean$readmit30 == "Yes", 1, 0), predicted = rf_pred_prob)
rf_calibration_plot_data <- rf_calibration_df %>% mutate(decile = ntile(predicted, 10)) %>% group_by(decile) %>% summarise(mean_predicted = mean(predicted), mean_observed = mean(observed), .groups = "drop")
brier_rf <- mean((rf_calibration_df$observed - rf_calibration_df$predicted)^2)
```

## Random Forest: ROC Curve

```{r rf-roc, fig.width=8, fig.height=5.5}
plot(roc_rf, col = "#33A02C", lwd = 3, main = "ROC Curve - Random Forest")
abline(a = 0, b = 1, lty = 2, col = "grey50")
```

## Random Forest: Calibration

```{r rf-calib, fig.width=8, fig.height=5}
ggplot(rf_calibration_plot_data, aes(x = mean_predicted, y = mean_observed)) +
  geom_point(size = 3, colour = "#33A02C") +
  geom_line(colour = "#33A02C") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey40") +
  labs(
    title = "Calibration Plot - Random Forest",
    x = "Mean Predicted Probability",
    y = "Observed Proportion"
  ) +
  theme_minimal()
```

## Random Forest: Variable Importance

```{r rf-varimp, fig.width=8, fig.height=5.5}
varImpPlot(rf_model, main = "Variable Importance - Random Forest")
```

**Interpretation**: Variables like `num_lab_procedures`, `num_medications`, `time_in_hospital`, and `age` consistently emerge as the most important predictors in the model. This emphasizes that healthcare utilization and demographic factors during the initial admission are key indicators of potential readmission risk.

## Model Comparison

```{r model-comp, fig.width=8, fig.height=5}
plot(roc_logit, col = "#1F78B4", lwd = 3, main = "ROC Curves for Both Models")
plot(roc_rf, col = "#33A02C", lwd = 3, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend(
  "bottomright",
  legend = c(
    paste("Logistic Regression AUC =", round(as.numeric(auc(roc_logit)), 3)),
    paste("Random Forest AUC =", round(as.numeric(auc(roc_rf)), 3))
  ),
  col = c("#1F78B4", "#33A02C"),
  lwd = 3,
  bty = "n"
)
```

## Summary & Conclusion

- **Logistic Regression**: 
  - AUC: `r round(as.numeric(auc(roc_logit)), 3)` | Brier Score: `r round(brier_logit, 4)`
  - Performance is slightly better than random baseline, limited discriminatory ability.
- **Random Forest**:
  - AUC: `r round(as.numeric(auc(roc_rf)), 3)` | Brier Score: `r round(brier_rf, 4)`
  - Outperformed Logistic Regression with better overall discrimination and prediction accuracy.
- **Conclusion**: Predicting 30-day hospital readmission is challenging; Random Forest identifies high-risk patients better.

## Project Information

- Full code and analysis available on GitHub:
- **[https://github.com/sararaees786/Predictive-Modelling](https://github.com/sararaees786/Predictive-Modelling)**
