---
title: "Prediction of 30-Day Hospital Readmission in Patients with Diabetes"
author: "Sara Raees"
output: html_document
date: "2026-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
required_packages <- c("tidyverse", "janitor", "caret", "naniar", "pROC", "randomForest", "skimr", "scales")

installed <- rownames(installed.packages())

for (pkg in required_packages) {
  if (!(pkg %in% installed)) {
    install.packages(pkg, dependencies = TRUE)
  }
}

library(tidyverse)
library(janitor)
library(skimr)
library(caret)
library(naniar)
library(pROC)
library(randomForest)
library(dplyr)
library(scales)
```

# Background

This project uses the Diabetes 130-US Hospitals for Years 1999–2008 dataset from the UCI Machine Learning Repository. The aim is to develop a prediction model for 30-day hospital readmission among hospitalized patients with diabetes.

For this assignment, a smaller stratified sample of the full dataset is used to reduce computational burden while preserving the proportion of patients readmitted within 30 days.

# Load, Read and Inspect the dataset

```{r}
diabetes_full <- read.csv("diabetic_data.csv", stringsAsFactors = FALSE)

# Create a smaller stratified sample for the assignment
set.seed(123)
sample_index <- createDataPartition(diabetes_full$readmitted, p = 20000 / nrow(diabetes_full), list = FALSE)
diabetes <- diabetes_full[sample_index, ]

names(diabetes)
str(diabetes)
dim(diabetes)

table(diabetes$readmitted)

diabetes[diabetes == "?"] <- NA

diabetes$readmit30 <- ifelse(diabetes$readmitted == "<30", 1, 0)
diabetes$readmit30 <- factor(diabetes$readmit30, levels = c(0,1), labels = c("No", "Yes"))

table(diabetes$readmit30)

missing_summary <- sapply(diabetes, function(x) sum(is.na(x)))
missing_table <- data.frame(
  Variable = names(missing_summary),
  Missing_Count = as.numeric(missing_summary),
  Missing_Percent = round(as.numeric(missing_summary) / nrow(diabetes) * 100, 2)
)
missing_table <- missing_table[order(-missing_table$Missing_Count), ]
head(missing_table, 15)
```

```{r}
model_data <- diabetes %>%
  select(readmit30, race, gender, age, admission_type_id,
         discharge_disposition_id, admission_source_id,
         time_in_hospital, num_lab_procedures, num_procedures,
         num_medications, number_outpatient, number_emergency,
         number_inpatient, number_diagnoses, max_glu_serum,
         A1Cresult, insulin, change, diabetesMed)

str(model_data)
summary(model_data)
```

# Initial data overview

```{r}
glimpse(diabetes)
summary(diabetes)
```

# Replace "?" with NA

```{r}
diabetes <- diabetes %>%
  mutate(across(where(is.character), ~ trimws(.))) %>%
  mutate(across(where(is.character), ~ na_if(., "?")))
```

# Define the outcome variable

The original variable `readmitted` has three categories:

- `"NO"` = no readmission  
- `">30"` = readmission after 30 days  
- `"<30"` = readmission within 30 days  

We create a binary outcome:

- `1` = readmitted within 30 days  
- `0` = not readmitted within 30 days  

```{r}
diabetes <- diabetes %>%
  mutate(
    readmit30 = ifelse(readmitted == "<30", 1, 0),
    readmit30 = factor(readmit30, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

# Study population and outcome frequency

```{r}
# Total number of observations
nrow(diabetes)

# Total number of predictors before excluding IDs/outcome
ncol(diabetes) - 1

# Outcome distribution
table(diabetes$readmitted)
table(diabetes$readmit30)

# Proportion readmitted within 30 days
prop.table(table(diabetes$readmit30))
```

# Count number of patients readmitted within 30 days

```{r}
sum(diabetes$readmitted == "<30", na.rm = TRUE)
```

# Missing data assessment

```{r}
# Number of missing values per variable
missing_counts <- colSums(is.na(diabetes))

# Proportion missing per variable
missing_prop <- colMeans(is.na(diabetes)) * 100

missing_summary <- data.frame(
  Variable = names(diabetes),
  Missing_Count = missing_counts,
  Missing_Percent = round(missing_prop, 2)
) %>%
  arrange(desc(Missing_Percent))

missing_summary
```

# Visualize missingness
```{r}
diabetes %>%
  slice_sample(n = min(5000, nrow(diabetes))) %>%
  vis_miss()
```

# Descriptive analysis of selected variables

```{r}
# Age distribution
table(diabetes$age)

# Gender distribution
table(diabetes$gender, useNA = "ifany")

# Race distribution
table(diabetes$race, useNA = "ifany")

# Summary of selected numeric variables
diabetes %>%
  select(time_in_hospital, num_lab_procedures, num_procedures,
         num_medications, number_outpatient, number_emergency,
         number_inpatient, number_diagnoses) %>%
  summary()
```

```{r}
# Coloured visualisations
ggplot(diabetes, aes(x = age, fill = readmit30)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("No" = "#4DAF4A", "Yes" = "#E41A1C")) +
  labs(
    title = "Proportion of 30-Day Readmission by Age Group",
    x = "Age Group",
    y = "Proportion",
    fill = "Readmitted in 30 Days"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(diabetes, aes(x = readmit30, y = time_in_hospital, fill = readmit30)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("No" = "#377EB8", "Yes" = "#FF7F00")) +
  labs(
    title = "Time in Hospital by 30-Day Readmission Status",
    x = "30-Day Readmission",
    y = "Time in Hospital"
  ) +
  theme_minimal()
```

```{r}
ggplot(diabetes, aes(x = num_medications, fill = readmit30)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("No" = "#984EA3", "Yes" = "#E41A1C")) +
  labs(
    title = "Distribution of Number of Medications",
    x = "Number of Medications",
    y = "Count",
    fill = "Readmitted in 30 Days"
  ) +
  theme_minimal()
```

# Select variables for modelling

Remove identifier variables that should not be used as predictors.

```{r}
diabetes_model <- diabetes %>%
  select(-encounter_id, -patient_nbr, -readmitted)
```

# Convert character variables to factors

```{r}
diabetes_model <- diabetes_model %>%
  mutate(across(where(is.character), as.factor))
```

# Remove variables with very high missingness

```{r}
missing_threshold <- 0.5

diabetes_model <- diabetes_model %>%
  select(where(~ mean(is.na(.)) < missing_threshold))

dim(diabetes_model)
```

# Train/test split

```{r}
set.seed(123)

train_index <- createDataPartition(diabetes_model$readmit30, p = 0.7, list = FALSE)

train_data <- diabetes_model[train_index, ]
test_data  <- diabetes_model[-train_index, ]

dim(train_data)
dim(test_data)
```

# Simple imputation

For a first analysis:

- numeric variables: median imputation  
- categorical variables: create `"Missing"` level  

```{r}
train_data_imputed <- train_data
test_data_imputed  <- test_data

# Numeric median imputation
num_vars <- names(train_data_imputed)[sapply(train_data_imputed, is.numeric)]

for (v in num_vars) {
  med <- median(train_data_imputed[[v]], na.rm = TRUE)
  if (is.na(med)) med <- 0
  train_data_imputed[[v]][is.na(train_data_imputed[[v]])] <- med
  test_data_imputed[[v]][is.na(test_data_imputed[[v]])] <- med
}

# Categorical missing category
cat_vars <- names(train_data_imputed)[sapply(train_data_imputed, is.factor)]

for (v in cat_vars) {
  train_data_imputed[[v]] <- addNA(train_data_imputed[[v]])
  levels(train_data_imputed[[v]])[is.na(levels(train_data_imputed[[v]]))] <- "Missing"

  test_data_imputed[[v]] <- factor(test_data_imputed[[v]], levels = levels(train_data_imputed[[v]]))
  test_data_imputed[[v]] <- addNA(test_data_imputed[[v]])
  levels(test_data_imputed[[v]])[is.na(levels(test_data_imputed[[v]]))] <- "Missing"
}
```

# Logistic regression model

```{r}
# Convert character to factor
train_data_imputed <- train_data_imputed %>%
  mutate(across(where(is.character), as.factor))

test_data_imputed <- test_data_imputed %>%
  mutate(across(where(is.character), as.factor))

# Match test levels to train
for (v in names(train_data_imputed)) {
  if (is.factor(train_data_imputed[[v]]) && v %in% names(test_data_imputed)) {
    test_data_imputed[[v]] <- factor(test_data_imputed[[v]],
                                     levels = levels(train_data_imputed[[v]]))
  }
}
```

```{r}
# Drop unused levels
train_data_imputed <- train_data_imputed %>%
  mutate(across(where(is.factor), droplevels))

test_data_imputed <- test_data_imputed %>%
  mutate(across(where(is.factor), droplevels))

# Remove factor columns with fewer than 2 observed categories
bad_vars <- names(train_data_imputed)[sapply(train_data_imputed, function(x) {
  if (is.factor(x)) {
    length(unique(na.omit(x))) < 2
  } else {
    FALSE
  }
})]

bad_vars

train_data_imputed <- train_data_imputed %>% select(-all_of(bad_vars))
test_data_imputed  <- test_data_imputed %>% select(-all_of(bad_vars))

# Remove near-zero variance predictors
nzv <- nearZeroVar(train_data_imputed)
if (length(nzv) > 0) {
  train_data_imputed <- train_data_imputed[, -nzv, drop = FALSE]
  test_data_imputed  <- test_data_imputed[, colnames(train_data_imputed), drop = FALSE]
}
```

```{r}
# Fit logistic regression
logit_model <- glm(readmit30 ~ ., data = train_data_imputed, family = binomial)
summary(logit_model)
```

# ROC and AUC

```{r}
test_data_imputed$readmit30 <- factor(test_data_imputed$readmit30, levels = c("No", "Yes"))

# Fix unseen factor levels in test data to match training data
factor_cols_logit <- names(train_data_imputed)[sapply(train_data_imputed, is.factor)]

for (col in factor_cols_logit) {
  if (col %in% names(test_data_imputed)) {
    test_data_imputed[[col]] <- as.character(test_data_imputed[[col]])
    test_data_imputed[[col]][!(test_data_imputed[[col]] %in% levels(train_data_imputed[[col]]))] <- NA
    test_data_imputed[[col]] <- factor(test_data_imputed[[col]], levels = levels(train_data_imputed[[col]]))
  }
}

# Keep only complete cases for prediction
test_logit_clean <- test_data_imputed[complete.cases(test_data_imputed), ]

# Logistic regression predicted probabilities
logit_pred_prob <- predict(logit_model, newdata = test_logit_clean, type = "response")

# ROC curve
library(pROC)

roc_logit <- roc(response = test_logit_clean$readmit30,
                 predictor = logit_pred_prob,
                 levels = c("No", "Yes"),
                 direction = "<")

# AUC
auc(roc_logit)

# Plot ROC
plot(roc_logit, col = "#1F78B4", lwd = 3,
     main = "ROC Curve - Logistic Regression")
abline(a = 0, b = 1, lty = 2, col = "grey50")
```


# Create calibration data
```{r}
calibration_df <- data.frame(
  observed = ifelse(test_logit_clean$readmit30 == "Yes", 1, 0),
  predicted = logit_pred_prob
)

# Group predictions into deciles
calibration_plot_data <- calibration_df %>%
  mutate(decile = ntile(predicted, 10)) %>%
  group_by(decile) %>%
  summarise(
    mean_predicted = mean(predicted),
    mean_observed = mean(observed),
    .groups = "drop"
  )

calibration_plot_data
```

# Calibration plot
```{r}
ggplot(calibration_plot_data, aes(x = mean_predicted, y = mean_observed)) +
  geom_point(size = 3, colour = "#E41A1C") +
  geom_line(colour = "#E41A1C") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey40") +
  labs(
    title = "Calibration Plot - Logistic Regression",
    x = "Mean Predicted Probability",
    y = "Observed Proportion"
  ) +
  theme_minimal()
```

# Brier score
```{r}
brier_logit <- mean((calibration_df$observed - calibration_df$predicted)^2)
brier_logit
```


#Random Forest Model
```{r}
# Remove high-cardinality factor predictors from training data
high_card_vars <- names(train_data_imputed)[sapply(train_data_imputed, function(x) {
  is.factor(x) && nlevels(x) > 53
})]

high_card_vars

train_rf <- train_data_imputed %>%
  select(-all_of(high_card_vars))

test_rf <- test_data_imputed %>%
  select(all_of(names(train_rf)))

# Ensure outcome levels are consistent
train_rf$readmit30 <- factor(train_rf$readmit30, levels = c("No", "Yes"))
test_rf$readmit30  <- factor(test_rf$readmit30,  levels = c("No", "Yes"))
```

```{r}
set.seed(123)

rf_model <- randomForest(
  readmit30 ~ .,
  data = train_rf,
  ntree = 300,
  importance = TRUE
)

print(rf_model)
```

# Test set predictions

```{r}
# Test set predictions

# Ensure outcome is a factor with consistent levels
train_rf$readmit30 <- factor(train_rf$readmit30, levels = c("No", "Yes"))
test_rf$readmit30  <- factor(test_rf$readmit30, levels = c("No", "Yes"))

# Keep test set columns in same order as training set
test_rf <- test_rf[, names(train_rf), drop = FALSE]

# Fix factor levels in test set to match training set
factor_cols <- names(train_rf)[sapply(train_rf, is.factor)]

for (col in factor_cols) {
  if (col %in% names(test_rf)) {
    test_rf[[col]] <- as.character(test_rf[[col]])
    
    # Replace unseen levels with NA
    test_rf[[col]][!(test_rf[[col]] %in% levels(train_rf[[col]]))] <- NA
    
    # Convert back to factor with training levels
    test_rf[[col]] <- factor(test_rf[[col]], levels = levels(train_rf[[col]]))
  }
}

# Match numeric/integer/logical types as well
for (col in names(train_rf)) {
  if (col %in% names(test_rf)) {
    if (is.numeric(train_rf[[col]]) && !is.factor(train_rf[[col]])) {
      test_rf[[col]] <- as.numeric(test_rf[[col]])
    } else if (is.integer(train_rf[[col]])) {
      test_rf[[col]] <- as.integer(test_rf[[col]])
    } else if (is.logical(train_rf[[col]])) {
      test_rf[[col]] <- as.logical(test_rf[[col]])
    }
  }
}

# Remove rows with NA values created by unseen factor levels
test_rf_clean <- test_rf[complete.cases(test_rf), ]

# Random forest predictions
rf_pred_class <- predict(rf_model, newdata = test_rf_clean, type = "class")
rf_pred_prob  <- predict(rf_model, newdata = test_rf_clean, type = "prob")[, "Yes"]

# Confusion matrix
library(caret)
confusionMatrix(data = rf_pred_class,
                reference = test_rf_clean$readmit30,
                positive = "Yes")
```


# ROC and AUC for random forest
```{r}
roc_rf <- roc(
  response = test_rf_clean$readmit30,
  predictor = rf_pred_prob,
  levels = c("No", "Yes"),
  direction = "<"
)

auc(roc_rf)

plot(roc_rf, col = "#33A02C", lwd = 3,
     main = "ROC Curve - Random Forest")
abline(a = 0, b = 1, lty = 2, col = "grey50")
```
# Calibration for random forest
```{r}
rf_calibration_df <- data.frame(
  observed = ifelse(test_rf_clean$readmit30 == "Yes", 1, 0),
  predicted = rf_pred_prob
)

rf_calibration_plot_data <- rf_calibration_df %>%
  mutate(decile = ntile(predicted, 10)) %>%
  group_by(decile) %>%
  summarise(
    mean_predicted = mean(predicted),
    mean_observed = mean(observed),
    .groups = "drop"
  )

rf_calibration_plot_data
```

```{r}
ggplot(rf_calibration_plot_data, aes(x = mean_predicted, y = mean_observed)) +
  geom_point(size = 3, colour = "#33A02C") +
  geom_line(colour = "#33A02C") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey40") +
  labs(
    title = "Calibration Plot - Random Forest",
    x = "Mean Predicted Probability",
    y = "Observed Proportion"
  ) +
  theme_minimal()
```
# Brier score for random forest
```{r}
brier_rf <- mean((rf_calibration_df$observed - rf_calibration_df$predicted)^2)
brier_rf
```

# Variable importance plot
```{r}
varImpPlot(rf_model, main = "Variable Importance - Random Forest")
```

# Compare model performance
```{r}
model_comparison <- data.frame(
  Model = c("Logistic Regression", "Random Forest"),
  AUC = c(as.numeric(auc(roc_logit)), as.numeric(auc(roc_rf))),
  Brier_Score = c(brier_logit, brier_rf)
)

model_comparison
```

# Combined ROC plot
```{r}
plot(roc_logit, col = "#1F78B4", lwd = 3,
     main = "ROC Curves for Both Models")
plot(roc_rf, col = "#33A02C", lwd = 3, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend(
  "bottomright",
  legend = c(
    paste("Logistic Regression AUC =", round(as.numeric(auc(roc_logit)), 3)),
    paste("Random Forest AUC =", round(as.numeric(auc(roc_rf)), 3))
  ),
  col = c("#1F78B4", "#33A02C"),
  lwd = 3,
  bty = "n"
)
```
# Brief Summary
```{r}
cat("Total observations:", nrow(diabetes), "\n")
cat("Total variables:", ncol(diabetes), "\n")
cat("Number readmitted within 30 days:",
    sum(diabetes$readmitted == "<30", na.rm = TRUE), "\n")
cat("Proportion readmitted within 30 days:",
    round(mean(diabetes$readmitted == "<30", na.rm = TRUE) * 100, 2), "%\n")
cat("Logistic Regression AUC:", round(as.numeric(auc(roc_logit)), 3), "\n")
cat("Random Forest AUC:", round(as.numeric(auc(roc_rf)), 3), "\n")
cat("Logistic Regression Brier Score:", round(brier_logit, 4), "\n")
cat("Random Forest Brier Score:", round(brier_rf, 4), "\n")
```

The dataset contains 20,001 observations and 51 variables. Among these, 2,232 patients were readmitted within 30 days, giving a readmission proportion of 11.16%. This indicates that the outcome is relatively uncommon, with most patients not being readmitted within 30 days.

The logistic regression model produced an AUC of 0.506, which is very close to 0.5. This suggests that the model has almost no discriminatory ability and performs only slightly better than random guessing in distinguishing between patients who were and were not readmitted within 30 days. Its Brier score was 0.155, which also indicates relatively weak predictive accuracy.

In contrast, the random forest model achieved an AUC of 0.619, showing better discrimination than logistic regression. Although this is still only a modest level of predictive performance, it suggests that the random forest model is more effective at identifying patients at higher risk of 30-day readmission. The Brier score for random forest was 0.0979, which is notably lower than that of logistic regression, indicating better overall prediction accuracy and calibration.

Overall, the random forest model performed better than logistic regression on this dataset, both in terms of discrimination and prediction error. However, the AUC of 0.619 suggests that prediction of 30-day readmission remains challenging, and further model improvement or inclusion of more informative predictors may be needed.
